Supermartingales and Termination Analysis of Statistical PCF

### 2020-08-31 probabilisticProgrammingMartingales.tex
Write the proof of thm:AstEquivalence

\begin{lemma}
For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
\end{lemma}
\begin{proof}
Suppose, to begin with, that $n = 0$. Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.

@@

\begin{theorem} \label{thm:AstEquivalence}
A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
\end{theorem}
\begin{proof}
For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length. The distributions are equal for each length, by induction, as follows.

@@

For the inductive case, let the distribution of terms after $n$ steps be $D_n$. This can be split into a random and a deterministic part, $D_n^r + D_n^d$, where $D_n^r$ has weight only on those terms of the form $E[\tsample]$, and $D_n^d$ only has weight on the terms not of that form. 
\end{proof}

>>>>

\begin{lemma} \label{lem:independentSamples}
For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
\end{lemma}
\begin{proof}
Suppose, to begin with, that $n = 0$. Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.

@@

\begin{theorem} \label{thm:AstEquivalence}
A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
\end{theorem}
\begin{proof}
\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}

For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length. The distributions are equal for each length, by induction, as follows.

@@

For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal. \akr{TODO: mention the case that it has already terminated by step $n$, or is about to. The two versions of the semantics handle it a little differently.}

For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$. The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments. If $R \new \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$. The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.

In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$. If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample. For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent. The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$. Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$. In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.

In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_\cbv$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.

The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
\end{proof}
